\textit{A young boy with a tree made of trees}: \\ Domain Adaptation of an LSTM-based Image Caption Generation Model

### 2023-01-09 paper/paper.tex
Final proofreading of the paper

    Within this project we aim to explore how fine-tuning a CNN and LSTM-based image captioning model on a new dataset influences its performance on both the old and new data, along with what role some of the parameters, such as fine-tuning data size or the prevalence of out-of-vocabulary tokens in captions play in the process. We discover that differences in vocabulary play a major role and that providing too much fine-tuning data can cause the whole system to fail. We conclude that domain adaptation can be difficult to carry out, but perhaps worth attempting in certain cases.

@@

In Section 2 we describe the materials and methods used to address the aforementioned questions, such as the architecture of the model, the datasets, and the evaluation methods. In Section 3 we present the results of our inquiry. In Section 4 we discuss how they answer our questions and how they relate to previous work. Finally, in Section 5 we offer conclusions and suggestions for future projects on this topic.\footnote{(MS)}

@@

Within this section we will describe the resources that we used in the project. We will specify the characteristics of the image captioning model, the two datasets that were used to train or fine-tune it, and the questionnaire we utilized to elicit human judgements of the best captions.\footnote{(MS)}

@@

The goal of this project was to analyze how the generation of captions changes, when an already pre-trained model is fine-tuned on new data from a different domain. For this, we utilized code written by \cite{ilinykh} that implemented a model described by \cite{xu2015attend}. We didn't adapt the model at all, but only fine-tuned it with a different dataset. In the following paragraph, the architecture model is summarized.

@@

\textit{Write about the division into files, a bit what's what, what we added and what was there. I did the evaluation part. I doubt we need to describe it in much detail, can always say that the readme has instructions and all the elements are documented, but I'd still have a bit of a summary of what we did here.}

Loading the data was done in a \emph{PyTorch Dataset} class. Here, it is possible to define multiple parameters for loading the data. First of all, only captions are loded that contain relations. Here, we implemented two different methods. One is filtering the captions based on the POS and the other is a rule-based approach taken from \cite{Ghanimifard2018}. Furthermore, a total number of images that should be loaded can be defined. This loaded samples are then divided into training (0.8), validation (0.1) and testing (0.1) sets. Since the imageCLEF captions consists of multiple connected captions, we included an option to either just load the first caption or concatenate all with the conjunction \emph{and}. Lastly, we noted that changing the dataset presents us with a lot of words that are not included in the original vocabulary. To deal with this, we implemented a variable filter threshold to exclude captions that contain too many unknown words (\emph{unknown filter}).

The training loop was mostly taken from \cite{ilinykh}. The pre-trained model is hereby fine-tuned on an unseen shuffled train dataset. After each epoch the model is validated with a validation dataset. This is done using BLEU-4 score. Since the imageCLEF dataset only consists of one independent caption instead of five as the Flickr8k dataset, the BLEU score is calculated only with one reference caption. The model is trained as long as the BLEU score doesn't improve for 20 epochs; the model with the best BLEU score is saved and used for testing and evaluating.

@@

Other than relatively minor fixes that were required to make the code run, we decided to manually exclude the UNK token from the generation process by enforcing the lowest possible probability on it in the beam search. While this meant that we could not evaluate to what extent the models favored generating that token, it did allow us to obtain captions that sounded more natural, which was paramount for the subsequent human evaluation.\footnote{(MS)}

@@

\textit{Here we should summarize our results, in terms of:
    \begin{enumerate}
        \item where did the performance exceed the best prior performance (we got BEST checkpoints?) \textbf{DK}
        \item which models actually generated reasonable captions? where did they start falling apart? \textbf{MS or DK}
    \end{enumerate}
    Include figures and tables as much as we can/as much as is reasonable.}

@@

During the training, we tried different hyperparameters for loading the data. Changing the method for filtering captions that contain relations didn't provide big differences in first tests. Therefore, we utilized solely the simpler POS-based approach for all experiments.

Concatenating all captions of the imageCLEF dataset was a too big challenge for the model and yielded worse results in first test than only using the first caption for each image. We concluded that the difference in syntax (more complex and longer sentences) as well as in semantics (more detailed descriptions) were too big to the Flickr8k dataset, the model was pre-trained on, but this hypothesis would need to be tested further in the future. For our experiments, we only used the first caption.

For the experiments, the variable parameters were the \emph{number of samples} as well as the \emph{unknown filter}. The model was fine-tuned on a sample size (training set size) of 100 (80), 200 (160), 500 (400), 1,000 (800), 2,000 (1,600) and 5,000 (4,000). For each of these sample sizes, four different unknown filter thresholds were used: 0.1, 0.15, 0.2 and 1.

As described previously in the Materials and Methods section, the only models that consistently produced captions for almost all test images were the original and the "100" models. The "200" models still generated most of the captions, but many of them started being rather nonsensical or non-grammatical. This process continued in the "500" and "1000" models with fewer and fewer captions being generated successfully, and those generated ones being absolutely nonsensical. Finally, in the "2k" and "5k" models most captions could not be generated.\footnote{MS}

While the human evaluation is the focus of this experiment, it is still insightful to also check, how the model fares during training and validation, more specifically, how the BLEU score changes. Interestingly, the models with few data samples achieve very low BLEU scores, while the models with more samples achieve higher BLEU scores. Table \ref{BLEU-sample-sizes} shows the best achieved BLEU scores during validation with an unknown filter of 0.1. For other filter values, the results look similar. This is opposite to the observation mentioned before. A reason for this may be the calculation of the BLEU score. The BLEU score only evaluates the appearance of words, in this experiment 4-grams, but does not focus explicitly on grammaticality. A lower BLEU score, but grammatical correct sentences may mean that the model doesn't learn to relate the new captions to the image features, but that it still relies on the pre-trained generations. On the other hand, a higher BLEU score with low grammaticality may hint to the fact, that the model is now able to generate the correct words, but looses the pre-trained ability to understand syntax.

@@

Looking at the unknown filter, higher thresholds produce in general better BLEU scores. Table \ref{BLEU-unknown-filter} shows the highest BLEU score for different unknown filter with 100 samples. This applies especially to the models with a larger number of samples, but the effect can be seen in all models. A reason for this may be the higher number of unknown tokens in the training data. Since, the model is also trained on generating unknown tokens, a higher share of unknown tokens in the training data will result in a higher share in the generated caption. Both sentences are therefore getting more similar, which results in a higher BLEU score. A solution for this problem could be excluding the unknown token from the model's vocabulary.

@@

While the captions themselves can be seen in the Jupyter Notebook files, the aim of the questionnaire was to reveal which model or models performed the best on which test set. Therefore, the results presented here are divided between the sets and include only the information on which model's caption was deemed the best, as well as how many people voted for it. Table 1 contains the information for the unfiltered imageCLEF set, Table 2 - for the filtered one, and Table 3 for Flickr8k. Image names are the file names without the .jpg extension. In some cases more than one model's caption received the same amount of votes, or the winning caption was generated by more than one model.

@@

As can be noticed in Table 1, the majority of the "good" captions for the unfiltered imageCLEF set were generated by the original, non-fine-tuned model, or by the one with a 10\% filter. Some captions from the model fine-tuned on non-filtered data turned out to have been rated high as well. Finally, other models' captions were rated as good when they overlapped with the original model's caption. Overall, while the judgements are split rather evenly, the original model seems to generate better captions for this test set. It is worth pointing out that relatively few captions were unanimously voted best.

When it comes to the filtered imageCLEF test set (Table 2), where a maximum of 10\% tokens in the original caption could be unknown tokens, the spectrum of what models' captions were considered good is even wider. There was only one case where one caption was voted best by all the participants. Half of the best captions were generated by the original model, and half by various fine-tuned ones, with none of them being a clear front-runner in this case.

Finally, when it comes to the original Flickr8k-based test set, as seen in Table 3, the original model without fine-tuning performed best, although there were instances where the 10\% or 20\% fine-tuned models generated better captions. Many of the original model's captions were also chosen unanimously.

@@

As seen in this project, domain adaption for a pretrained image captioning model can be very challenging. It was surprising, to which extent the pre-trained model was disrupted when fine-tuned on data from another domain. Even though the dataset for fine-tuning was very different than the dataset for pre-training, we hoped that the fine-tuned model would still perform better on the new domain than the original. This was not the case for multiple reasons, mentioned above. At least, we expected it to be able to generate sensible captions for most of the images and that it would only produce big mistakes for a smaller part. With the rather opposite results discussed before, two major challenges for domain adaption got highlighted. First, the model can't adapt well to a different structure of captions. This results in ungrammatical sentences, when fine-tuned on many samples from the new domain. Secondly, the model is not able to adapt to a different topic of images. That results in 'reusing' old captions on new images and not learning its image features, when fine-tuned on few samples. For a deeper analysis, it may be helpful to first fine-tune the model on a domain that is closer in one of these aspects, for example a domain, that uses a similar way of describing the images, but focuses on topics. This could test the hypothesis made in this paper.

@@

Throughout this project we have explored and tested how a CNN and LSTM-based image caption generation model could be fine-tuned, or adapted to a new domain, as represented by a somewhat different image-caption dataset. We have identified points that can be especially problematic, such as differences in the vocabulary and the subsequent generation of UNK tokens in the captions. We have determined that if the domains are too different and too much fine-tuning data is fed to the model, the generation will start to completely fail or focus only on a very narrow range of words, producing very ungrammatical captions. We have also collected human judgements and determined that while fine-tuning does slightly increase the quality of the captions on the new dataset, all of them, regardless of the model, were still considered weird, mismatched, or unnatural, and that the structure of the sentences in the original dataset may have a large influence on how they are generated even after fine-tuning (e.g. trying to feature people in every caption).

While we have addressed the issues that we raised in the introduction, we were not able to explore all the possible variables in domain adaptation of models of this kind, as it was beyond the scope of this project. However, we are certain that it would be very interesting to see how other hyperparameters than just the training data size can influence the outcome, and what the interplay between them could be. It would also be fascinating to see similar research pertaining to other model architectures, especially trying to identify the optimal hyperparameters for fine-tuning transformer-based image caption generation models. Finally, it would be interesting to explore other potential ways of dealing with the vocabulary mismatch between different datasets.

@@

\textit{Here we have our statement of who did what.}
While really detailed information on who developed what is available in the code, the code is not the only thing that we shared work on. Below one can find the details of who contributed what to the project.
\begin{enumerate}
    \item \textbf{Developing project ideas:} both Dominik and Maria. Selecting the initial ideas from the ones suggested in the course page mostly by Dominik, ideas about working with comicbooks and contacting researchers related to that by Maria. Final idea selected under the advice from Nikolai and Simon by both project authors.
    \item \textbf{Setting up resources:} figuring out how to access the datasets used in the project by Dominik, using the existing code by Nikolai both by Maria and Dominik.
    \item \textbf{Code:} detailed information on code contribution can be found in code documentation.
          \begin{enumerate}
              \item Evaluation Jupyter Notebooks: Maria
              \item Dataset classes: Dominik
              \item Relation filters: Maria and code shared by Simon
              \item Fine-tuning loop: adapted by Maria and Dominik from Nikolai's code.
              \item Models and some preprocessing: Nikolai's code.
              \item Caption generation and attention visualization: Nikolai's code, with changes by Maria and Dominik.
              \item Division of the code into separate files, turning it into classes: Dominik.
              \item Documentation: Maria with Dominik's help.
          \end{enumerate}
    \item \textbf{Questionnaire:} Maria.
    \item \textbf{Evaluation:} evaluation of the outputs of the notebooks by Dominik and Maria (for the questionnaire), evaluation of questionnaire answers by Maria.
    \item \textbf{Writing up:} both Dominik and Maria. Abstract by Maria (cannot put the initials there). While the initials signify the person who wrote most of the section, small changes by the other person are possible. Some sections' authorship is shared; in that case a footnote is put at the end of the part written by one person.
    \item \textbf{Repository management:} Maria with Dominik's help.
\end{enumerate}

>>>>

    Within this project we aim to explore how fine-tuning a CNN and LSTM-based image captioning model on a new dataset influences its performance on both the old and new data, along with what role some of the parameters, such as fine-tuning data size or the prevalence of out-of-vocabulary tokens in captions play in the process. We discover that differences in vocabulary play a major role and that providing too much fine-tuning data can cause the whole system to fail. We conclude that domain adaptation can be difficult to carry out, but perhaps worth attempting in certain cases (MS).

@@

In Section 2 we describe the materials and methods used to address the aforementioned questions, such as the architecture of the model, the datasets, and the evaluation methods. In Section 3 we present the results of our inquiry. In Section 4 we discuss how they answer our questions and how they relate to previous work. Finally, in Section 5 we offer conclusions and suggestions for future projects on this topic.

@@

Within this section we will describe the resources that we used in the project. We will specify the characteristics of the image captioning model, the two datasets that were used to train or fine-tune it, and the questionnaire we utilized to elicit human judgements of the best captions (MS)

@@

The goal of this project was to analyze how the generation of captions changes when an already pre-trained model is fine-tuned on new data from a different domain. For this, we utilized code written by \cite{ilinykh} that implemented a model described by \cite{xu2015attend}. We didn't adapt the model at all, but only fine-tuned it with a different dataset. In the following paragraph, the architecture model is summarized.

@@

Loading the data was done in a \emph{PyTorch Dataset} class. Here, it is possible to define multiple parameters for loading the data. First of all, only captions are loded that contain relations. Here, we implemented two different methods. One is filtering the captions based on the POS tag, and the other is a rule-based approach taken from \cite{Ghanimifard2018}. Furthermore, a total number of images that should be loaded can be defined. This loaded samples are then divided into training (0.8), validation (0.1) and testing (0.1) sets. Since the imageCLEF captions consists of multiple connected captions, we included an option to either just load the first caption or concatenate all with the conjunction \emph{and}. Lastly, we noted that changing the dataset presents us with a lot of words that are not included in the original vocabulary. To deal with this, we implemented a variable filter threshold to exclude captions that contain too many unknown words (\emph{unknown filter}).

The training loop was mostly taken from \cite{ilinykh}. The pre-trained model is hereby fine-tuned on an unseen shuffled train dataset. After each epoch the model is validated with a validation dataset. This is done using BLEU-4 score. Since the imageCLEF dataset only consists of one independent caption instead of five as the Flickr8k dataset, the BLEU score is calculated only with one reference caption. The model is trained as long as the BLEU score doesn't improve for 20 epochs; the model with the best BLEU score is saved and used for testing and evaluating (DK).

@@

Other than relatively minor fixes that were required to make the code run, we decided to manually exclude the UNK token from the generation process by enforcing the lowest possible probability on it in the beam search. While this meant that we could not evaluate to what extent the models favored generating that token, it did allow us to obtain captions that sounded more natural, which was paramount for the subsequent human evaluation (MS).

@@

@@

During the training, we tried different hyperparameters for loading the data. Changing the method for filtering captions that contain relations didn't provide big differences in first tests. Therefore, we utilized solely the simpler POS-based approach that we developed ourselves for all experiments.

Concatenating all captions of the imageCLEF dataset was a too big of a challenge for the model and yielded worse results in first test than only using the first caption for each image. We concluded that the difference in syntax (more complex and longer sentences) as well as in semantics (more detailed descriptions) were too big to the Flickr8k dataset, the model was pre-trained on, but this hypothesis would need to be tested further in the future. For our experiments, we only used the first section of the caption.

For the experiments, the variable parameters were the \emph{number of samples} as well as the \emph{unknown filter}. The model was fine-tuned on a sample size (training set size) of 100 (80), 200 (160), 500 (400), 1,000 (800), 2,000 (1,600) and 5,000 (4,000). For each of these sample sizes, four different unknown filter thresholds were used: 0.1, 0.15, 0.2 and 1 (DK).

As mentioned previously in the Materials and Methods section, the only models that consistently produced captions for almost all test images were the original and the "100" models. The "200" models still generated most of the captions, but many of them started being rather nonsensical or non-grammatical. This process continued in the "500" and "1000" models with fewer and fewer captions being generated successfully, and those generated ones being absolutely nonsensical. Finally, in the "2k" and "5k" models most captions could not be generated (MS).

While the human evaluation is the focus of this experiment, it is still insightful to also check, how the model fares during training and validation, more specifically, how the BLEU score changes. Interestingly, the models with few data samples achieve very low BLEU scores, while the models with more samples achieve higher BLEU scores. Table \ref{BLEU-sample-sizes} shows the best achieved BLEU scores during validation with an unknown filter of 0.1. For other filter values, the results look similar. This is opposite to the observation mentioned before. A reason for this may be the calculation of the BLEU score. The BLEU score only evaluates the appearance of words, in this experiment 4-grams, but does not focus explicitly on grammaticality. A lower BLEU score, but grammatical correct sentences may mean that the model does not learn to relate the new captions to the image features, but that it still relies on the pre-trained generations. On the other hand, a higher BLEU score with low grammaticality may hint to the fact, that the model is now able to generate the correct words, but looses the pre-trained ability to understand syntax.

@@

Looking at the unknown filter, higher thresholds produce in general better BLEU scores. Table \ref{BLEU-unknown-filter} shows the highest BLEU score for different unknown filter with 100 samples. This applies especially to the models with a larger number of samples, but the effect can be seen in all models. A reason for this may be the higher number of unknown tokens in the training data. Since, the model is also trained on generating unknown tokens, a higher share of unknown tokens in the training data will result in a higher share in the generated caption. Both sentences are therefore getting more similar, which results in a higher BLEU score. A solution for this problem could be excluding the unknown token from the model's vocabulary (DK).

@@

While the captions themselves can be seen in the Jupyter Notebook files, the aim of the questionnaire was to reveal which model or models performed the best on which test set. Therefore, the results presented here are divided between the sets and include only the information on which model's caption was deemed the best, as well as how many people voted for it. Table \ref{unfiltered-table} contains the information for the unfiltered imageCLEF set, Table \ref{filtered-table} - for the filtered one, and Table \ref{flickr-table} for Flickr8k. Image names are the file names without the .jpg extension. In some cases more than one model's caption received the same amount of votes, or the winning caption was generated by more than one model.

@@

As can be noticed in Table \ref{unfiltered-table}, the majority of the "good" captions for the unfiltered imageCLEF set were generated by the original, non-fine-tuned model, or by the one with a 10\% filter. Some captions from the model fine-tuned on non-filtered data turned out to have been rated high as well. Finally, other models' captions were rated as good when they overlapped with the original model's caption. Overall, while the judgements are split rather evenly, the original model seems to generate better captions for this test set. It is worth pointing out that relatively few captions were unanimously voted best.

When it comes to the filtered imageCLEF test set (Table \ref{filtered-table}), where a maximum of 10\% tokens in the original caption could be unknown tokens, the spectrum of what models' captions were considered good is even wider. There was only one case where one caption was voted best by all the participants. Half of the best captions were generated by the original model, and half by various fine-tuned ones, with none of them being a clear front-runner in this case.

Finally, as for the original Flickr8k-based test set, as seen in Table \ref{flickr-table}, the original model without fine-tuning performed best, although there were instances where the 10\% or 20\% fine-tuned models generated better captions. Many of the original model's captions were also chosen unanimously.

@@

As seen in this project, domain adaption for a pretrained image captioning model can be very challenging. It was surprising to what extent the pre-trained model was disrupted when fine-tuned on data from another domain. Even though the dataset for fine-tuning was very different than the dataset for pre-training, we hoped that the fine-tuned model would still perform better on the new domain than the original. This was not the case for multiple reasons, mentioned above. At the very least, we expected it to be able to generate sensible captions for most of the images and that it would only produce big mistakes for a smaller part. With the rather opposite results discussed before, two major challenges for domain adaption got highlighted. First, the model cannot adapt well to a different structure of captions. This results in ungrammatical sentences, when fine-tuned on many samples from the new domain. Secondly, the model is not able to adapt to a different topic of images. That results in 'reusing' old captions on new images and not learning its image features, when fine-tuned on few samples. For a deeper analysis, it may be helpful to first fine-tune the model on a domain that is closer in one of these aspects, for example a domain, that uses a similar way of describing the images, but focuses on topics. This could test the hypothesis made in this paper.

@@

Throughout this project we have explored and tested how a CNN and LSTM-based image caption generation model could be fine-tuned, or adapted to a new domain, as represented by a somewhat different image-caption dataset. We have identified points that can be especially problematic, such as differences in the vocabulary and sentence structure, and the subsequent generation of UNK tokens in the captions as well as ungrammatical sentences. We have determined that if the domains are too different and too much fine-tuning data is fed to the model, the generation will start to completely fail or focus only on a very narrow range of words. We have also collected human judgements and determined that while fine-tuning does slightly increase the quality of the captions on the new dataset, all of them, regardless of the model, were still considered weird, mismatched, or unnatural, and that the structure of the sentences in the original dataset may have a large influence on how they are generated even after fine-tuning (e.g. trying to feature people in every caption).

While we have addressed the issues that we raised in the introduction, we were not able to explore all the possible variables in domain adaptation of models of this kind, as it was beyond the scope of this project. However, we are certain that it would be very interesting to see how other hyperparameters than just the training data size can influence the outcome, and what the interplay between them could be. It would also be fascinating to see similar research pertaining to other model architectures, especially trying to identify the optimal hyperparameters for fine-tuning transformer-based image caption generation models. Of course, following up on the idea from the Discussion section, to verify our conclusions from the results by fine-tuning on a more similar dataset is another topic worth investigating. Finally, it would be interesting to explore other potential ways of dealing with the vocabulary mismatch between different datasets.

@@

While really detailed information on who developed what is available in the code, the code is not the only thing that we shared work on. Below one can find the details of who contributed what to the project.
\begin{enumerate}
    \item \textbf{Developing project ideas:} both Dominik and Maria. Selecting the initial ideas from the ones suggested in the course page mostly by Dominik, ideas about working with comicbooks and contacting researchers related to that by Maria. Final idea selected under the advice from Nikolai and Simon by both project authors.
    \item \textbf{Setting up resources:} figuring out how to access the datasets used in the project by Dominik, using the existing code by Nikolai both by Maria and Dominik.
    \item \textbf{Code:} detailed information on code contribution can be found in code documentation.
          \begin{enumerate}
              \item Evaluation Jupyter Notebooks: Maria
              \item Dataset classes: Dominik
              \item Relation filters: Maria and code shared by Simon
              \item Fine-tuning loop: adapted by Maria and Dominik from Nikolai's code.
              \item Models and some preprocessing: Nikolai's code.
              \item Caption generation and attention visualization: Nikolai's code, with changes by Maria and Dominik.
              \item Division of the code into separate files, turning it into classes: Dominik.
              \item Documentation: Maria with Dominik's help.
          \end{enumerate}
    \item \textbf{Questionnaire:} Maria.
    \item \textbf{Evaluation:} evaluation of the outputs of the notebooks by Dominik and Maria (for the questionnaire), evaluation of questionnaire answers by Maria.
    \item \textbf{Writing up:} both Dominik and Maria. While the initials signify the person who wrote most of the section, small changes by the other person are possible - and are marked by (initials) at the end of a paragraph. This is done especially if a certain paragraph was moved from one section to the other. Otherwise, for sections where authorship is more evenly shared, the same method of marking contribution is employed.
    \item \textbf{Repository management:} Maria with Dominik's help.
    \item \textbf{Presentation:} Maria with Dominik's help.
\end{enumerate}
